### 2020-09-16 paper/experiments.tex
Fixes

Let us describe experiments and their results. All of our algorithms search for an odd hole or an odd antihole and stop when find one, or a evidence of one. Therefore, their running times are greatest with perfect graphs on the input and we decided to use mainly\todo{run some smart nonperfect tests} perfect graphs for our performance benchmarks. For every test, the vertex number were shuffled. For every randomly generated graph class and for each size, we ran the tests on 10 graphs generated with different seeds and took their average running time.

@@

We generate random bipartite graphs by creating two equal-sized sets of vertices and every possible edge between them has a probability of $p = 0.5$ to appear. In this way we can also generate line graphs of bipartite graphs. This gives us graphs of varying $|V|$, so we repeated the process until we had sufficieng number of graphs for each $|V|$.

There are also a few interesting classes of perfect graphs that appear on a grid, or a checker board. In lattice graph, each vertex is connected to vertices that are above, below, to the left and to the right of it, if such vertices exist. See \Cref{fig:lattice} for an example of a $4 \times 4$ lattice graph. If we imagine our grid to be a chessboard we can define a rook graph to have an edge, when the rook from chess can move from one vertex to another in one move. See \Cref{fig:rook} for an example of $4 \times 4$ rook graph. In similar manner we can define a knight graph (\Cref{fig:knight}).

@@

Split graphs we use are unions of cliques and independent sets of same size, with some edges between them. In generating them, each edge between a vertex from a clique and a vertex from an independent set has had $1/2$ chance to appear.

@@

@@

First, let us take a look at the performance of CCLSV algorithm. If we limit runtime to about two minutes, we can run it on graphs of sizes up to 40-70, depending on the type of the graph we test. What is interesting, that there isn't that much of a variance of the running time depending on the type of graph we test on. This is a big advantage of CCLSV over the na\"ive algorithm, which is highly dependent on the type of input. On some graphs, it works almost instantenously, what is to be expected, given its nature of enumerating paths. But on some graphs, e.g. lattice graphs (\Cref{plot:gridLines}) its overall running time grows much faster than CCLSV's.

@@

Examples of full binary graphs, and lattice graphs show that getting near cleaners still can be a big bottleneck. Parallelization of this method could be considered in the future. A possible idea would be to use recently developed dynamic dictionary for the GPU \cite{Ashkiani2018}. When running GPU CCLSV on split graphs, quite a lot of time is spent checking simple structures. This should be parallelizable, the biggest challange we can see is spreading work evenly across CUDA threads.

@@

\subsection{Ellipsoid method}

@@

Calculating $\vartheta(G)$ is the most complicated part of the coloring algorithm. With that done by an external library, the rest of the program is a straightforward implementation of the algorithms in \Cref{sec:coloringEllipsoid}. In most of our tests, the majority of running time was consumed on calculating $\vartheta$ of various graphs. As it was done by an external library, there isn't much optimization potential for us. One thing of note is, that specifying the precision of $\vartheta$ we want to calculate to be $1/3$ sped up the algorithm by TODO. Also, we cache results of calculatin $\vartheta$, meaning that throught whole execution of our coloring algorithm, we only calculate $\vartheta$ for each graph once. Our main goal of the implementation was to check if this method is still impractical, even on modern equipment. Let us proceed straight to experiments and results.

@@

First, we present overall running times of the coloring algorithm (\Cref{fig:CSDPLines}) and then breakdown the running time in two parts: calculating $\vartheta$ by the CSDP library and graphs manipulations which constitute the rest of the coloring algorithm (\Cref{fig:CSDPDet}). We note that $\vartheta$ calculation is highly dependent on the type of the graph. For our biggest tested cases of $|V| = 48$, all required calculations of $\vartheta$ take almost 7 minutes in biparite graphs, but only 12 seconds in line graphs of bipartite graphs. Because CSDP library is highly optimized, it is expected that it would run much faster on some graphs. However, it is not clear what could we do to speedup the running times on tests that CSDP is slow on, such as bipartite graphs of lattice graphs. A potential improvement would be to again use CUDA, this time for $\vartheta$ calculations, possibly utilizing the cuSOLVER library \cite{cusolver}.

We notice that on some graphs, namely line graphs of bipartite graphs, rook graphs and split graphs the time taken by operations other than calculating $\vartheta$ by the CSDP library is quite significant. However, the overall running times on these graphs are low, so we didn't particularly optimize this, so probably some improvements could be made here.

\section{Conclusions}
We implemented and benchmarked two different approaches for recognizing perfect graphs. Both have their positives and drawbacks -- CCLSV's running times are more predictable, but on many graphs it is noticeably slower than the na\"ive algorithm. On the other hand, running na\"ive algorithm on a graph we don't know any properties of can lead to unbearably slow running time.

@@

We also implemented algorithm for coloring perfect graphs. It is not entirely unusable, although far from fast. Maybe a better way of calculating $\vartheta$ could be found, or the GPU could be again usable for this task. We point to Nvidia's \cite{cusolver} as a possible starting point.

>>>>

Let us describe experiments and their results. All of our algorithms search for an odd hole or an odd antihole and stop when find one, or a evidence of one. Therefore, their running times are greatest with perfect graphs on the input and we decided to use\todo{run some smart nonperfect tests} perfect graphs for our performance benchmarks. For every test, the vertex number were shuffled. For every randomly generated graph class and for each size, we ran the tests on 10 graphs generated with different seeds and took their average running time.

@@

We generate random bipartite graphs by creating two equal-sized sets of vertices, and then added every possible edge between them with a probability of $p = 0.5$. In this way we can also generate line graphs of random bipartite graphs. This gives us graphs of varying $|V|$, so we repeated the process until we had sufficieng number of graphs for each $|V|$.

There are also a few interesting classes of perfect graphs that appear on a grid, or a checker board. In lattice graph, each vertex is connected to vertices that are above, below, to the left and to the right of it, if such vertices exist. See \Cref{fig:lattice} for an example of a $4 \times 4$ lattice graph. If we imagine our grid to be a chessboard we can define two vertices a rook graph to share an edge, when the rook from chess can move from one vertex to the other in one move. See \Cref{fig:rook} for an example of $4 \times 4$ rook graph. In similar manner we can define a knight graph (\Cref{fig:knight}).

@@

Split graphs we use are unions of cliques and independent sets of same size, with some edges between them. In generating them, each edge between a vertex from a clique and a vertex from an independent set has had a chance $p = 0.5$ to appear.

@@

We ran our tests on Intel Xeon E5-2690 CPU and Nvidia Tesla K80 GPU with 4992 CUDA cores.

@@

First, let us take a look at the performance of CCLSV algorithm. If we limit runtime to about two minutes, we can run it on graphs of sizes up to 40-70, depending on the type of the graph we test.  Worth noting is the fact that there isn't that much of a variance of the running time depending on the type of graph we test on. This is a big advantage of CCLSV over the na\"ive algorithm, which is highly dependent on the type of input. On some graphs, it works almost instantenously, what is to be expected, given its method of enumerating paths. But on some graphs, e.g. lattice graphs (\Cref{plot:gridLines}) its overall running time grows much faster than CCLSV's.

@@

Examples of full binary graphs, and lattice graphs show that getting possible near cleaners still can be a big bottleneck. Parallelization of this method could be considered in the future. A possible idea would be to use recently developed dynamic dictionary for the GPU \cite{Ashkiani2018}. When running GPU CCLSV on split graphs, quite a lot of time is spent checking simple structures. The biggest challange we can see is in an attempt to parallelise this is spreading work evenly across CUDA threads.

@@

\subsubsection{Ellipsoid method}

@@

Calculating $\vartheta(G)$ is the most complicated part of the coloring algorithm. With that done by an external library, the rest of the program is a straightforward implementation of the algorithms in \Cref{sec:coloringEllipsoid}. In most of our tests, the majority of running time was consumed on calculating $\vartheta$ of various graphs. As it was done by an external library, there isn't much optimization potential for us. One thing of note is, that by specifying the precision of $\vartheta$ we want to calculate to be $1/3$ the algorithm sped up  by 20-40\%. Also, we cache results of calculatin $\vartheta$, meaning that throught whole execution of our coloring algorithm, we only calculate $\vartheta$ for each graph once. Our main goal of the implementation was to check if this method is still impractical, even on modern equipment. Let us proceed straight to experiments and results.

@@

First, we present overall running times of the coloring algorithm (\Cref{fig:CSDPLines}) and then breakdown the running time in two parts: calculating $\vartheta$ by the CSDP library and graphs manipulations which constitute the rest of the coloring algorithm (\Cref{fig:CSDPDet}). We note that $\vartheta$ calculation is highly dependent on the type of the graph. For our biggest tested cases of $|V| = 48$, all required calculations of $\vartheta$ take almost 7 minutes in bipartite graphs, but only 12 seconds in line graphs of bipartite graphs. Because CSDP library is highly optimized, it is expected that it would run much faster on some graphs. However, it is not clear what could we do to speedup the running times on tests that CSDP is slow on, such as bipartite graphs or lattice graphs. A potential improvement would be to again use CUDA, this time for $\vartheta$ calculations, possibly utilizing the cuSOLVER library \cite{cusolver}.

We notice that on some graphs, namely line graphs of bipartite graphs, rook graphs and split graphs the time taken by operations other than calculating $\vartheta$ is quite significant. However, the overall running times on these graphs are low, so we didn't particularly optimize this, so probably some improvements could be made here.

\section{Conclusions}
We implemented and benchmarked two different approaches for recognizing perfect graphs. Both have their positives and drawbacks -- CCLSV's running times are more predictable, but on many graphs it is noticeably slower than the na\"ive algorithm. On the other hand, running na\"ive algorithm on a graph we don't know any properties of can lead to unbearably slow performance.

@@

We also implemented algorithm for coloring perfect graphs. It is not entirely unusable, although far from fast. Maybe a better way of calculating $\vartheta$ could be found, or the GPU could again be usable for this task. We point to Nvidia's cuSOLVER  cite{cusolver} as a possible starting point.

### 2020-09-16 paper/implementation.tex
Fixes

\section{Berge graphs recognition: na\"ive approach}
When implementing an algorithm with the running time of $O(n^9)$ (especially one as complex as CCLSV), one of the first questions that pop up is whether this algorithm is applicable at all. We couldn't find any existing programs that test whether the graph is perfect, be it an implementation of CCLSV, or a na\"ive approach.\todo{well.. there's this java impl} So, to test the usability of CCLSV implementation, we measured it against our own na\"ive perfect recognition algorithm. Let us describe it briefly.

We try find an odd hole directly, then, we run the same algorithm on $\overline{G}$, and if no odd hole is found we report that the graph is Perfect, else that it is not.

To find an odd hole, we use a backtracking technique. We will enumerate all paths and holes and for each test if it is an odd hole. A partial solution to our backtracking problem is a path in $G$. To move forward from a partial solution, we append consecutively to its end all neighbors of current path's last vertex and continue recursively after each. When a current solution is an odd hole we return it. When we exhaust all possible paths from the current partial solution, we remove last vertex and continue.

This backtracking algorithm is very simple, but has a great potential for optimizations in implementation. We also use path enumeration algorithm in the CCLSV algorithm, so we give much attention to its optimization.

@@

This simple change in data structure design gave us a speedup of overall running time of our algorithm in the range of about 5x.

Another crucial optimization is to eliminate partial solutions that have edges between not consecutive vertices right away. Also, when we look for an odd hole specifically, we can assume that its first vertex we enumerate is the lowest. Those two optimizations make a huge difference. With them, our na\"ive algorithm implementation went from being able to check graphs with $|V|$ up to 12-14, within a couple of minutes, to being faster than CCLSV on most of our tests.

@@

With all those optimizations made, our na\"ive algorithm proved to be quite fast in some cases, although its running time is very dependent on the properties of the input, as we should expect from an optimized non-polynomial algorithm. See \Cref{sec:experiments} for running times.

@@

The Berge recognition algorithm's running time is $O(n^9)$, which brings into question its applicability to any real use case. Although time complexity is indeed a limiting factor, a number of lower level optimizations done on implementation's level make a very big difference and make it more viable than na\"ive algorithm, at least on some test cases. We also explore a new frontier of implementing its most time consuming part on massively parallel GPU architecture, with some good results (\Cref{sec:CUDA}).

@@

When implementing a complicated algorithm that has a time complexity of $O(n^9)$ optimizations can be both crucial and difficult to implement. There is not a single code path that takes up all the running time -- or at lest there isn't one from a theoretical point of view. Therefore a tool for inspecting running time bottlenecks is needed. We used Valgrind's tool called \emph{callgrind} \cite{callgrind}.

@@

Callgrind is a profiling tool that records the call history and event counts ((data reads, cache misses etc.) of a program and presents it as a call-graph. It is then possible to visualise this, we used a tool \emph{gprof2dot} \cite{gprof2dot} to generate visual call graphs from callgrind's output.

Before any optimizatios, a major part of Berge recognition algorithm was being spent on enumerating all possible paths of given length -- this is done either to find a hole by itself, find a simple structure or check a possible near cleaner for amenable odd hole. Therefore optimizations of the path enumerating algorithm were also helpful in speeding up CCLSV implementation.

@@

Another algorithm for which we found major speedups is the algorithm to generate all possible near-cleaners \Cref{alg:listNearCleaners}). We optimized it by using a a \texttt{dynamic\_bitset} from the boost library \cite{boost}. It is a data structure to represent a set of bits, that also allows for fast bitwise operators that one can apply to builtin integers. At the very end of \Cref{alg:listNearCleaners} a set $\mathcal{R}$ is constructed, which is an union of all possible unions of pairs $X_i$ and $N_j$. We used \texttt{dynamic\_bitset} to represent each of the sets $X_i$ and $N_j$, and the set $\mathcal{R}$. This significantly speeds up the operation of union of two sets (line \ref{line:listNCcalcR}). With this modification, the speedup of \Cref{alg:listNearCleaners} was about 20\%.

@@

Also, we test the algorithm on graphs that we know are perfect -- such as bipartite graphs, line graphs of bipartite graphs and complements thereof. In addition to that, a test on all perfect graphs up to a size of 10 was performed -- we used \cite{graphRepo} as a source of perfect graphs.

We also test on graphs that we know are not perfect: we generate them by generating an odd hole and adding it with some edges to a random graph.

@@

CUDA is a parrallel computing platform model created by Nvidia and executed on Nvidia GPUs. It allows to run a program in by many CUDA threads at once. Although usually described by the the PRAM models, implementing PRAM algorithm in CUDA architecture is not as straightforward as implementing a RAM algorith on a CPU. Some of the causes are: synchronization between threads is very costly, there are multiple memory models to choose from, threads run divided in blocks and it is possible to efficiently share memory within blocks and copying memory from and back to CPU suffers from a large latency. What is more, there are no standard dynamic structures in the PRAM model, let alone CUDA, that are taken for granted in the RAM model, such as hashing maps or sets. 
% We used a library called \emph{moderngpu} for our implementation, because it allows a slightly cleaner code.

@@

Let us recall the \Cref{alg:testNearCleaner}. It calculates array $R$ of shortest paths, then for each 3-vertex path and an additional vertex it does $O(1)$ checks to see if they along with two paths from $R$ give us an odd hole. We will parallelise the work after calculating $R$, that is, lines \ref{line:cudaStart}-\ref{line:cudaEnd}. Let us notice, that for all $X$s all 3-vertex paths enumerated in line \ref{line:3vertex} are the same. We calculate them beforehand, then each CUDA thread receives one 3-vertex path $x_1 - x_2 - x_3$ and an additional vertex $y_1$ and performs the required checks in lines \ref{line:cudaWorkStard}-\ref{line:cudaWorkEnd}. This is almost perfect scenario for the GPU -- we do a simple SIMD work in parallel, without a lot scattered memory access.

The part of the CCLSV implemented on the GPU achieved speedup of up to 15x. Which gave us a speedup of up to 9x for overall CCLSV algorithm.

>>>>

\section{Berge graphs recognition: na\"ive approach}
When implementing an algorithm with the running time of $O(|V|^9)$ (especially one as complex as CCLSV), one of the first questions that pop up is whether this algorithm is at all usable. We couldn't find any existing programs that test whether the graph is perfect, be it an implementation of CCLSV, or a na\"ive approach.\todo{well.. there's this java impl} So, to test the usability of CCLSV implementation, we measured it against our own na\"ive perfect recognition algorithm. Let us describe it briefly.

We try find an odd hole directly, then, we run the same algorithm on $\overline{G}$, and if no odd hole is found we report that the graph is perfect, else that it is not.

To find an odd hole, we use a backtracking technique. We will enumerate all paths and holes, and for each test if it is an odd hole. A partial solution to our backtracking problem is a path in $G$. To move forward from a partial solution, we append consecutively to its end all neighbors of current path's last vertex and continue recursively after each. When a current solution is an odd hole we return it. When we exhaust all possible paths from the current partial solution, we remove its last vertex and continue.

This backtracking algorithm is very simple, but has a great potential for optimizations in its implementation. We also use path enumeration algorithm in the CCLSV algorithm, so we give much attention to its optimization.

@@

This simple change in data structure design gave us a speedup of overall running time of our na\"ive algorithm in the range of about 5x.

Another crucial optimization is to eliminate partial solutions that have edges between not consecutive vertices right away. Also, when we look for an odd hole specifically, we can assume that its first vertex we enumerate is the lowest. Those two optimizations make a huge difference. With them, our na\"ive algorithm implementation went from being able to check graphs with $|V|$ up to 12-14, within a reasonable time, to being faster than CCLSV on most of our tests.

@@

With all those optimizations made, our na\"ive algorithm proved to be quite fast in some cases, although its running time is very dependent on the properties of the input, as we should expect from an optimized non-polynomial algorithm. See \Cref{sec:experiments} for the running times.

@@

The Berge recognition algorithm's running time is $O(|V|^9)$, which brings into question its applicability to any real use case. Although time complexity is indeed a limiting factor, a number of lower level optimizations done on implementation's level make a very big difference and make it more viable than na\"ive algorithm, at least on some test cases. We also explore a new frontier of implementing its most time consuming part on massively parallel GPU architecture, with some good results (\Cref{sec:CUDA}).

@@

When implementing a complicated algorithm that has a time complexity of $O(|V|^9)$ optimizations can be both crucial and difficult to implement. There is not a single code path that takes up all the running time -- or at least there isn't one from a theoretical point of view. Therefore a tool for inspecting running time bottlenecks is needed. We used Valgrind's tool called \emph{callgrind} \cite{callgrind}.

@@

Callgrind is a profiling tool that records the call history and event counts (data reads, cache misses etc.) of a program and presents it as a call-graph. It is then possible to visualise this, we used a tool \emph{gprof2dot} \cite{gprof2dot} to generate visual call graphs from callgrind's output.

Before any optimizatios, a major part of Berge recognition algorithm was being spent on enumerating all possible paths of given length -- this is done either to find a simple structure or check a possible near cleaner for an amenable odd hole. Therefore optimizations of the path enumerating algorithm were also helpful in speeding up CCLSV implementation.

@@

Another algorithm for which we found major speedups is the algorithm to generate all possible near-cleaners (\Cref{alg:listNearCleaners}). We optimized it by using a a \texttt{dynamic\_bitset} from the boost library \cite{boost}. It is a data structure to represent a set of bits, that also allows for fast bitwise operators that one can apply to builtin integers. At the very end of \Cref{alg:listNearCleaners} a set $\mathcal{R}$ is constructed, which is an union of all possible unions of pairs $X_i$ and $N_j$. We used \texttt{dynamic\_bitset} to represent each of the sets $X_i$ and $N_j$, and the elements of the set $\mathcal{R}$. This significantly speeds up the operation of union of two sets (line \ref{line:listNCcalcR}). With this modification, the speedup of \Cref{alg:listNearCleaners} was about 20\%.

@@

Also, we test our implementations on graphs that we know are perfect -- such as bipartite graphs, line graphs of bipartite graphs and complements thereof. In addition to that, a test on all perfect graphs up to a size of 10 was performed -- we used \cite{graphRepo} as a source of perfect graphs.

We also test on graphs that we know are not perfect: we generate them by generating an odd hole and joining it with some edges to a random graph.

@@

CUDA is a parrallel computing platform model created by Nvidia and executed on Nvidia GPUs. It allows to run a program  by many CUDA threads at once. Although usually described by the the PRAM models, implementing PRAM algorithm in CUDA architecture is not as straightforward as implementing a RAM algorith on a CPU. Some of the causes are: synchronization between threads is very costly, there are multiple memory models to choose from, threads run divided in blocks and it is possible to efficiently share memory within blocks and copying memory from and back to CPU suffers from a large latency. What is more, there are no standard dynamic structures in the PRAM model, let alone CUDA, that are taken for granted in the RAM model, such as hashing maps or sets. 
% We used a library called \emph{moderngpu} for our implementation, because it allows a slightly cleaner code.

@@

Let us recall the \Cref{alg:testNearCleaner}. It calculates array $R$ of shortest paths, then for each 3-vertex path and an additional vertex it does $O(1)$ checks to see if they along with two paths from $R$ give us an odd hole. We will parallelise the work after calculating $R$, that is, lines \ref{line:cudaStart}-\ref{line:cudaEnd}. Let us notice, that for all $X$s, all 3-vertex paths enumerated in line \ref{line:3vertex} are the same. We calculate them beforehand, then each CUDA thread receives one 3-vertex path $x_1 - x_2 - x_3$ and an additional vertex $y_1$ and performs the required checks in lines \ref{line:cudaWorkStard}-\ref{line:cudaWorkEnd}. This is almost perfect scenario for the GPU -- we do a simple SIMD work in parallel, without a lot scattered memory access.

The part of the CCLSV implemented on the GPU achieved speedup of up to 15x. Which gave us a speedup of up to 9x for overall CCLSV algorithm, depending on a class of graph we benchmark on.

### 2020-09-16 paper/main.tex
Poprawki up to old Ch 1
many many fixes
Fixes

\begin{abstract}
	TODOa
\end{abstract}

\listoftheorems[ignoreall,show={defn}]
\tableofcontents

@@

\chapter*{Definitions}
\input{definitions.tex}

\chapter{Perfect Graphs}
\input{perfectGraphs.tex}
\input{coloringBergeGraphs.tex}

\chapter{Implementation}
\input{implementation.tex}

>>>>

\begin{abstract}
	Perfect graphs are a subject of intense study since their introduction in 1961. Only in 2001 the famous Perfect graph conjecture was proven to be true and in 2005 a polynomial method of determining if a graph is perfect was found. Since 1988 an algorithm is known for coloring perfect graphs, but it uses an ellipsoid method which is said to be complicated and impractical. As recently as in 2018 a polynomial algorithm for coloring perfect graphs without squares was published, that uses a combinatorial approach.

	We begin with basic definitions in Chapter 1. In Chapter 2 we introduce perfect graphs and Berge graphs, give an overview of the strong perfect graph theorem and talk about an algorithm for polynomial perfect graph recognition. Chapter 3 is a study of the ellipsoid method of coloring perfect graphs. In Chapter 4 we present our implementation of algorithms from Chapters 2 and 3, along with notes on optimization and parallelisation. Appendix A is a overview of the recent algorithm for coloring square-free perfect graphs.
\end{abstract}

\listoftheorems[ignoreall,show={defn}]
\renewcommand{\listtheoremname}{List of algorithms}
\listoftheorems[ignoreall,show={alg}]
\tableofcontents

@@

\chapter{Basic Graph Definitions}
\input{definitions.tex}

\chapter{Perfect Graphs}
\input{perfectGraphs.tex}

\chapter{Coloring Perfect Graphs}
\input{coloringBergeGraphs.tex}

\chapter{Implementation}
\input{implementation.tex}
\input{experiments.tex}
